A Computational Account of Default-Mode Function\\by Control Theory and Reinforcement Learning

### 2016-08-04 paper.tex
fixups

\subsection{Demystifying the free-energy principle}
In this section we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision making. This ideas were first developed by Hinton et al. in the early 90s in the form of the so-called  \textit{Helmholtz machine}. Theories like Friston's free-energy principle will then emerge as particular instanstances of this general framework, with particular design choises. For example the Friston theory assumes the brain uses a (problematic) wake-sleep algorithm to train the underlying Helmholtz machine, etc.
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{11cm}}
         \hline
         \textbf{variable}    & \textbf{short description}  \\ \hline
             $\langle E\rangle_p$ & Expectation (a.k.a average) of the random quantity $E$
         .r.t to the probability density $p$.\\ \hline
         $\mathcal H$ & The usual information-theoretic entropy operator.\\ \hline
         $D_{KL}(p||q)$ & The Kullback-Leibler divergence between the probability densities $p$ and $q$. \\ \hline
             $\zeta$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's hidden state $b$ and sensory inputs $s$, i.e $\zeta = (s, b).$ \\ \hline
             $\psi$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
             $p_G(.|\zeta)$ & Generative density for ...\\ \hline
         $p_R(.|\zeta)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
         $F_G(\zeta)$ & Helmholtz free-energy for the model $p_G$ of generating the observation $\zeta$.\\ \hline
         $F^R_G(\zeta)$ & Variational Helmholtz free-energy from $R$ to $G$.  Note that $F^G_G = F_G$.\\ \hline
  \end{tabular}
  \caption{Table of notations.}
\end{table}
\subsubsection{Helmholtz free-energy and the generative model}
Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
\begin{eqnarray}
  \begin{split}
    \text{generative surprise } &= -\log(p_G(\zeta)) = -\log(p_G(\zeta)) \times 1 = -\log(P_G(\zeta))\sum_{\psi}p_G(\psi |\zeta)\\
    &= -\sum_{\psi}p_G(\psi, \zeta)\log(p_G(\zeta))
    =-\sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi, \zeta)/p_G(\psi|\zeta))\\
    &= \sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi|\zeta))-\sum_{\psi}p_G(\psi |\zeta)\log  (p_G(\psi, \zeta))\\
    &= -\langle \log  (p_G(., \zeta)) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))\\
    &= \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta)) := F_G(\zeta)
  \end{split}
  \label{eq:helm}
\end{eqnarray}
where $E_G(\psi, \zeta)$ is the energy of an fictive thermodynamic system defined by setting
\begin{equation}
  E_G(\psi, \zeta) := -\log(p_G(\psi, \zeta)).
  \label{eq:gibbs}
\end{equation}
%% In the above, the variable $\psi$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
The last quantity $F_G(\zeta) := \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))$ in \eqref{eq:helm} can be is nothing other than the Helmholtz free-energy for the physical system correspond to the external world, prescribed by the generative density $p_G(\psi|\zeta)$ for the occupation of \textit{macrostates} $\psi$. Thus generative surprise and generative Helmholtz free-energy are different views on exactly the same object.

@@

Now, invoking \eqref{eq:gibbs} and applying Bayes rule, we get the Gibbs
distribution
\begin{equation}
  p_G(\psi|\zeta) = \frac{p_G(\psi|\zeta)}{p_G(\zeta)} = \frac{\exp(-E_G(\psi, \zeta))}{Z_G(\zeta)} = \frac{\exp(-E_G(\psi, \zeta))}{Z_G(\zeta)},
\end{equation}
where $Z_G(\zeta) := \log(p_G(\zeta)) = \sum_{\psi'}\exp(-E_G(\psi',\zeta))$, the normalizing partition function for the model \ref{eq:gibbs}.
Whence $\forall \psi$, $Z_G(\zeta) = p_G(\zeta) = \exp(-E_G(\psi, \zeta)) / p_G(\psi|\zeta)$, and so we have the invariance relation
\begin{equation}
  -\log(Z_G(\zeta)) = -\log(p_G(\zeta)) = F_G(\zeta) = E_G(\psi, \zeta) + \log(p_G(\psi|\zeta)).
  \end{equation}
Now, in the equation above, the LHS only depends on the generative model $G$ and the data point $s$: it doesn't depend on the hidden variable $\psi$, etc. So, taking expectations w.r.t an arbitrary density\footnote{The conditioning in $P_R(.|\zeta)$ is because this density is selected from a world in which the sensory inputs and internal brain state vector $\zeta$ is assumed already observed.} $p_R(.|\zeta)$ (the subscrit $R$ stands for ``recognition'', and the terminology will become clear in a moment) yields
\begin{equation}
  \begin{split}
    F_G(\zeta) &= -\log(Z_G(\zeta)) = \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} + \sum_{\psi}p_R(\psi|\zeta)\log(p_G(\psi|\zeta))\\
    &= \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} - \mathcal H(p_R(.|\zeta)) - \sum_{\psi}p_R(\psi|\zeta)\log(p_R(\psi|\zeta)/p_G(\psi|\zeta))\\
    &= F^R_G(\zeta) - D_{KL}(P_R(.|\zeta) || P_G(.|\zeta)),
  \end{split}
  \label{eq:fe}
\end{equation}
where $F^R_G(\zeta)$ is the \textit{variational} Helmholtz free-energy from $R$ to $G$ defined by
\begin{equation}
  F^R_G(\zeta) := \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} - \mathcal H(p_R(.|\zeta))
\end{equation}
and $D_{KL}(P_R(.|\zeta) || P_G(.|\zeta))$ is the Kullback-Leibler divergence between the $p_R(.|\zeta)$ and the generative density $p_G(.|\zeta)$. Note that $F^G_G = F_G$.

>>>>

\subsection{Demystifying the free-energy principle}
In this section we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in the form of the so-called  \textit{Helmholtz machine}. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For example the Friston theory assumes the brain uses a (problematic) wake-sleep algorithm to train the underlying Helmholtz machine, etc.
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{11cm}}
         \hline
         \textbf{variable}    & \textbf{short description}  \\ \hline
         $\langle E\rangle_p$ & Expectation (a.k.a average) of the
         random quantity $E$ w.r.t to the probability density $p$.\\ \hline
         $\mathcal H$ & The usual information-theoretic entropy operator.\\ \hline
         $D_{KL}(p||q)$ & The Kullback-Leibler divergence between the probability densities $p$ and $q$. \\ \hline
             $\zeta$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's hidden state $b$ and sensory inputs $s$, i.e $\zeta = (s, b).$ \\ \hline
             $\psi$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
             $p_G(.|\zeta)$ & Generative density for ...\\ \hline
         $p_R(.|\zeta)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
         $F_G(\zeta)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\zeta$. This measures the surprise incured upon observing $\zeta$ generated by the model $G$.\\ \hline
         $F^R_G(\zeta)$ & Variational Helmholtz free-energy from model $R$
         to $G$.  Note that $F^G_G = F_G$.\\ \hline
  \end{tabular}
  \caption{Table of notations.}
\end{table}
\subsubsection{Helmholtz free-energy and the generative model}
Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
\begin{eqnarray}
  \begin{split}
    \text{generative surprise } &= -\log(p_G(\zeta)) = -\log(p_G(\zeta)) \times 1 = -\log(P_G(\zeta))\sum_{\psi}p_G(\psi |\zeta)\\
    &= -\sum_{\psi}p_G(\psi, \zeta)\log(p_G(\zeta))
    =-\sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi, \zeta)/p_G(\psi|\zeta))\\
    &= \sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi|\zeta))-\sum_{\psi}p_G(\psi |\zeta)\log  (p_G(\psi, \zeta))\\
    &= -\langle \log  (p_G(., \zeta)) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))\\
    &= \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta)) := F_G(\zeta)
  \end{split}
  \label{eq:helm}
\end{eqnarray}
where $E_G(\psi, \zeta)$ is the energy of an fictive thermodynamic system defined by setting
\begin{equation}
  E_G(\psi, \zeta) := -\log(p_G(\psi, \zeta)).
  \label{eq:gibbs}
\end{equation}
%% In the above, the variable $\psi$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
The last quantity $F_G(\zeta) := \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))$ in \eqref{eq:helm} can be is nothing other than the Helmholtz free-energy for the physical system correspond to the external world, prescribed by the generative density $p_G(\psi|\zeta)$ for the occupation of \textit{macrostates} $\psi$. Thus generative surprise and generative Helmholtz free-energy are different views on exactly the same object.

@@

Now, invoking \eqref{eq:gibbs} and applying Bayes rule, we get the Gibbs
distribution
\begin{equation}
  p_G(\psi|\zeta) = \frac{p_G(\psi|\zeta)}{p_G(\zeta)} = \frac{\exp(-E_G(\psi, \zeta))}{Z_G(\zeta)} = \frac{\exp(-E_G(\psi, \zeta))}{Z_G(\zeta)},
\end{equation}
where $Z_G(\zeta) := \log(p_G(\zeta)) = \sum_{\psi'}\exp(-E_G(\psi',\zeta))$, the normalizing partition function for the model \ref{eq:gibbs}.
Whence $\forall \psi$, $Z_G(\zeta) = p_G(\zeta) = \exp(-E_G(\psi, \zeta)) / p_G(\psi|\zeta)$, and so we have the invariance relation
\begin{equation}
  -\log(Z_G(\zeta)) = -\log(p_G(\zeta)) = F_G(\zeta) = E_G(\psi, \zeta) + \log(p_G(\psi|\zeta)).
  \end{equation}
Now, in the equation above, the LHS only depends on the generative model $G$ and the data point $\zeta$: it doesn't depend on the hidden variable $\psi$, etc. So, taking expectations w.r.t an arbitrary density\footnote{The conditioning in $P_R(.|\zeta)$ is because this density is selected from a world in which the sensory inputs and internal brain state vector $\zeta$ is assumed already observed.} $p_R(.|\zeta)$ yields
\begin{equation}
  \begin{split}
    F_G(\zeta) &= -\log(Z_G(\zeta)) = \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} + \sum_{\psi}p_R(\psi|\zeta)\log(p_G(\psi|\zeta))\\
    &= \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} - \mathcal H(p_R(.|\zeta)) - \sum_{\psi}p_R(\psi|\zeta)\log(p_R(\psi|\zeta)/p_G(\psi|\zeta))\\
    &= F^R_G(\zeta) - D_{KL}(P_R(.|\zeta) || P_G(.|\zeta)),
  \end{split}
  \label{eq:fe}
\end{equation}
where $F^R_G(\zeta)$ is the \textit{variational} Helmholtz free-energy from $R$ to $G$ defined by
\begin{equation}
  F^R_G(\zeta) := \langle E_G(., \zeta)\rangle_{P_R(.|\zeta)} - \mathcal H(p_R(.|\zeta))
\end{equation}
and $D_{KL}(P_R(.|\zeta) || P_G(.|\zeta))$ is the Kullback-Leibler divergence between the $p_R(.|\zeta)$ and the generative density $p_G(.|\zeta)$. Note that $F^G_G = F_G$.
